### 2020-05-18 podc_NEW/figs/timeline.tex
New BA

 \documentclass[a4paper,anonymous,USenglish]{lipics-v2019}

 \usepackage[utf8]{inputenc}
 \usepackage{xspace}
 \usepackage{balance}
 \usepackage{amsmath,amsfonts,mathtools,amsthm}
 \usepackage{algorithmic}
 \usepackage{algorithm}

 \usepackage{balance}
 \usepackage{amsthm,amsmath,array,colortbl,graphicx,multirow}
 \usepackage{comment}
 \usepackage{balance}
 \usepackage{tikz}
 \usepackage{amsmath}
 \usetikzlibrary{decorations.pathreplacing} %
 \usepackage{algorithm}
 \usepackage[font={footnotesize}]{subcaption}
 \usepackage[font={footnotesize}]{caption}
 \usepackage{breakcites}
 \usepackage{booktabs}
 \usepackage{diagbox}
 \usepackage{xcolor}
 \usepackage{colortbl}
 \usepackage{cleveref}
 \usepackage{enumitem}

 \mathchardef\mhyphen="2D

 \title{Optimal Algorithms for Dynamic Balanced Graph Partitioning}

 \author{Maciej Pacut}{maciej.pacut@univie.ac.at}{Faculty of Computer Science, University of Vienna,Austria}{0000-0002-6379-1490}{}

% \author{Mahmoud Parham}{mahmoud.parham@univie.ac.at}{Faculty of Computer Science, University of Vienna, Austria}{0000-0002-6211-077X}{}

% \author{Stefan Schmid}{stefan\_schmid@univie.ac.at}{Faculty of Computer Science, University of Vienna, Austria}{}{}

 \authorrunning{M. Pacut, M. Parham, and S. Schmid}
 \Copyright{Maciej Pacut, Mahmoud Parham, and Stefan Schmid}
 \keywords{online algorithms, competitive analysis, distributed computing, graph partitioning, clustering, self-adjusting networks}

% \EventEditors{}
%\EventNoEds{3}
 \EventLongTitle{34th International Symposium on DIStributed Computing (DISC) 2020}
 \EventShortTitle{DISC 2020}
 \EventAcronym{DISC}
 \EventYear{2020}
 \EventDate{October 12--16, 2020}
 \EventLocation{Freiburg, Germany (virtual conference)}
 \EventLogo{}
 \SeriesVolume{}
 \ArticleNo{}

 \ccsdesc[500]{Theory of computation~Online algorithms}
 \ccsdesc[500]{Networks~Network algorithms}
 %\ccsdesc[300]{Computer systems organization~Cloud computing}
 \ccsdesc[300]{Computer systems organization~Distributed architectures}

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%&&
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%&&
 %  our macros start
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%&&
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%&&

 \newcommand{\OPT}{\textsc{OPT}\xspace}
 \newcommand{\ALG}{\textsc{ALG}\xspace}
 \newcommand{\PPL}{\textsc{PPL}\xspace}
 \newcommand{\OBRP}{BRP\xspace}
 \newcommand{\PPOBRP}{PP-BRP}
 \newcommand{\dist}{\textsc{dist}}
 \newcommand{\TAlg}{{\ensuremath{\textsf{ALG}_{3}}}\xspace}
  \newcommand{\RM}{\textsc{RM}\xspace} % rematching alg

 \newcommand{\Rep}{\textsc{Rep}}

 %\newtheorem{claim}{Claim}
 \newtheorem{fact}{Fact}
 \newtheorem{rem}{Remark}
 \newtheorem{observation}{Observation}
 \newtheorem{property}{Property}

 \DeclarePairedDelimiter\pair{(}{)}
 \DeclarePairedDelimiter\set{\{}{\}}

 \DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
 \DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

 \newcommand\mahmoud[1]{\color{orange}\textbf{Mahmoud: #1~}\color{black}}
 \newcommand\stefan[1]{\color{blue}\textbf{Stefan: #1}\color{black}}
 \newcommand\maciek[1]{\color{brown}\textbf{(Maciek: #1)}\color{black}}
 %\newcommand\mahmoud[1]{}
 %\newcommand\stefan[1]{}
 %\newcommand\maciek[1]{}

 \newcommand{\todo}[1]{\noindent\color{brown}{todo: #1}\color{black}}

 \begin{CCSXML}
	<ccs2012>
	<concept>
	<concept_id>10003033.10003068</concept_id>
	<concept_desc>Networks~Network algorithms</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10010520.10010521.10010537.10003100</concept_id>
	<concept_desc>Computer systems organization~Cloud computing</concept_desc>
	<concept_significance>300</concept_significance>
	</concept>
	<concept>
	<concept_id>10010520.10010521.10010537</concept_id>
	<concept_desc>Computer systems organization~Distributed architectures</concept_desc>
	<concept_significance>300</concept_significance>
	</concept>
	</ccs2012>
\end{CCSXML}

% for my small screen
%\setlength{\textwidth}{8cm}

\begin{document}
\maketitle

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a~significant amount of network traffic.
	By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), we can reduce the network load and  improve application performance. 
	However, the communication pattern of different applications is often unknown a~priori and may change over time, hence it needs to be learned in an~online manner.
	%
	This paper revisits the online 
	balanced partitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	that asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a~significantly improved deterministic
	lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a~scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide an asymptotically tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	For $k=3$, we contribute an asymptotically tight upper bound
	of $\Theta(\ell)$
	for the general model in which the
	communication pattern can change arbitrarily over time.
	We improve the result for $k=2$ by providing a~strictly $6$-competitive upper bound for the general model.
	In contrast to most prior work, our algorithms respect all capacity constraints and do not require resource augmentation.

\end{abstract}

%\renewcommand{\shortauthors}{M.~Pacut, M.~Parham, S.~Schmid}

\section{Introduction}

The popularity of data-centric, distributed applications has led to an explosive growth of network traffic, especially in data centers~\cite{roy2015inside,singh2015jupiter}.
The performance of these distributed applications often critically depends on the underlying network~\cite{mogul2012we}, and efficient operation of these networks is important.
At the same time, distributed systems are often highly virtualized today, and provide interesting new opportunities for resource optimization.
In particular, it has become possible to operate data centers in a~more demand-aware manner: 
by dynamically migrating nodes (e.g., virtual machines) which communicate frequently topologically closer to each other, network traffic can be reduced significantly.  
However, migrations entail overhead and should be used moderately.

This paper studies the algorithmic problem underlying such demand-aware
optimizations, aiming to strike a~balance between the benefits of migrations (e.g., reduced network load) and their costs.
In particular, we are interested in an online variant of the problem: since communication patterns can change over time, an online algorithm needs to react dynamically to new traffic patterns, and migrate nodes  accordingly.
Ideally, this algorithm should perform close to an optimal offline algorithm, without requiring any information about future traffic demands.

\pagebreak

This problem is known as the dynamic balanced graph partitioning problem and was introduced by Avin et al.~\cite{repartition-disc, sidma-arxiv} at DISC 2016. a~special variant of the general problem has later been studied by Henzinger et al.~\cite{sigmetrics19_partitioning} at SIGMETRICS 2019.
We refer to the latter as the learning model.

\subsection{Model}

We study two models in this paper: the \emph{general partitioning} model, and its subproblem, the \emph{learning} model.
In both models, we assume that communication patterns are not known to our algorithms at the beginning.
We measure the~quality of~presented algorithmic solutions by competitive analysis~\cite{borodin-book}, which is well-suited for problems that are online by their nature.
In the competitive analysis, the goal is to~optimize \emph{the competitive ratio} of a~given online algorithm: the ratio of its cost to the cost of~an~optimal offline algorithm that knows the entire input sequence in advance.
%We emphasize that all our algorithms have a~\emph{strict} competitive ratio (without an additive term).

\noindent
\textbf{General partitioning model.}
In the \emph{dynamic balanced graph partitioning} problem, we are given a~set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a~sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a~pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

An algorithm serves a~communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to these two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a~request,
an online algorithm may perform a~\emph{repartition},
%(i.e., \emph{reconfigure}).
i.e.,
it may move (``migrate'') some nodes into clusters different from their current clusters, while respecting the capacity of every cluster. 
Afterward, 
the algorithm serves the  request.
The cost of migrating a~node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $\ALG$,
its cost,
denoted by $\ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $\ALG$ while serving the sequence $\sigma$.

\noindent
\textbf{Learning model.}
We study a~\emph{learning} variant of dynamic balanced graph partitioning,
where the communication pattern is \emph{static}:
whether a~pair of  nodes ever communicate or not, 
is determined a~priori and is unknown to algorithms,
 and such pairs communicate forever.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al.~\cite{sigmetrics19_partitioning}, we assume that the communication graph admits a~\emph{perfect partition},
i.e., a~partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.
For the learning model, we  assume that the migration cost is $\alpha=1$.

\subsection{Related Work}

The two works closest to ours are by Avin et al. (on the general partitioning model)~\cite{repartition-disc, sidma-arxiv} and by Henzinger et al. (on the learning model)~\cite{sigmetrics19_partitioning}.
However, the focus of these papers is primarily on models with resource augmentation: the online algorithm can use slightly larger clusters than the offline algorithm.  
Avin et al. actually showed that their lower bound $\Omega(k)$ holds even for a~significant resource augmentation, and they provided an algorithm with the competitive ratio $O(k \log k)$ using the $(2+\epsilon)$-augmented cluster capacity.
Their ratio is independent of $\ell$, which is impossible without significant resource augmentation.

In contrast, we study the non-augmented setting, where the nodes need to be perfectly balanced  among the clusters.
This assumption is not only more realistic but also significantly more challenging, as it is related to hard problems such as integer partitioning~\cite{integer-partitions-book}.
In terms of results without augmentation, so far, it is only known that there exists an~$O(k^2 \cdot \ell^2)$-competitive algorithm~\cite{repartition-disc}; the best known lower bound is significantly lower, namely $\Omega(k)$.
For $k=2$, Avin et al.~\cite{repartition-disc} presented a~$7$-competitive algorithm with a~substantial ($\Omega(\ell^2)$) additive constant.

The problem has also been studied in a~weaker
model where the adversary can only sample
requests from a~fixed distribution~\cite{stochastic-ring}.

The static offline version of~the~partitioning~problem, i.e., a~problem variant where
migration is not allowed, where all requests are known in advance, and where
the goal is to find an assignment of $n$ nodes to $\ell$~physical machines, each of~capacity $n/\ell$, is known as the
\emph{$\ell$-balanced graph partitioning problem}. The problem is 
NP-complete, and cannot even be approximated within any finite factor unless P
= NP~\cite{AndRae06}.  The static
variant where $\ell = 2$ corresponds to the minimum bisection problem, which
is already NP-hard~\cite{GaJoSt76}, and 
the currently best approximation ratio is $O(\log n)$~\cite{SarVaz95,ArKaKa99,FeKrNi00,FeiKra02,KraFei06,Raec08}.

Our problem is further related to some classic online problems.
In particular, it is related to online paging~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00}, sometimes also referred to
as online caching, where requests for data items (nodes) arrive over time and
need to be served from a~cache of finite capacity, and where the number of
cache misses must be minimized. Classic problem variants usually boil down to
finding a~smart eviction strategy, such as Least Recently Used (LRU)~\cite{SleTar85}. In our
setting, requests can be served remotely (i.e.,~without fetching the
corresponding nodes to a~single physical machine). In this light, our model is more
reminiscent of caching models \emph{with
bypassing}~\cite{EpImLN11,EpImLN15,Irani02}.
A major difference between  these problems is that in the caching problems, each request involves a~single element of the universe, while in our model \emph{both} endpoints of a~communication request are subject to~optimization.
In this light, we can see our model as a~"symmetric" version of online paging.

Dynamic graph partitioning problems are generally fundamental in computer science, and arise in many different contexts~\cite{streaming-soda,streaming1}.

%More generally, the model is related to online
%caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00},
%see~\cite{repartition-disc} for a~discussion.
%The static offline version of~the~problem, called the
%\emph{$\ell$-balanced graph partitioning problem} is 
%NP-complete, and cannot even be approximated within %any finite factor unless P
%= NP~\cite{AndRae06}.

\subsection{Our Contributions}

This paper presents several new results on the dynamic graph partitioning problem  without augmentation.
For the learning model, we present a~lower bound of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
(that holds also in the general partitioning model).
The best known lower bound so far was $\Omega(k)$~\cite{repartition-disc} that holds only in the general partitioning model.
We complement this result with an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.

For the general partitioning model, we design
an asymptotically optimal,
$\Theta(\ell)$-competitive algorithm for $k=3$, improving the best known upper bound 
so far $O(\ell^2)$~\cite{repartition-disc}.
We further present a~strictly $6$-competitive algorithm for $k=2$ that improves upon the previous $7$-competitive algorithm with $O(\alpha\ell^2)$ additive constant.

All algorithms in this paper have a strict competitive ratio (i.e., without an additive term).
Table \ref{tab:overview} provides an overview of our contributions compared to prior work.

\begin{table*}[h]
	\centering
	\renewcommand{\arraystretch}{1.5}
	\begin{tabular}{>{\centering\arraybackslash}p{4.25cm}|>{\centering\arraybackslash}p{4.25cm}>{\centering\arraybackslash}p{4.25cm}}
		\rowcolor{gray!50}
		\textbf{Variant} & \textbf{ Lower bound} &\textbf{Upper bound}\\ \hline 
		\textbf{$k=2$}& 3\hspace{0.3cm}\cite{repartition-disc} & 6\hspace{0.3cm}(\S \ref{sec:k2}) \\ 
		\rowcolor{gray!25}
		\textbf{$k=3$}&  $\Omega(\ell)$ \hspace{0.3cm}(\S \ref{sec:lowerbound})& $O(\ell) $\hspace{0.3cm}(\S \ref{sec:k3})\\
		$k > 3$ & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k^2 \cdot \ell^2)$\hspace{0.1cm}~\cite{repartition-disc} \\
		\rowcolor{gray!25}
		Learning model & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell)$\hspace{0.1cm} (\S \ref{sec:ppl}) \\
	\end{tabular}
	\caption{Overview of known results and our contributions. The table summarizes the results for the general partitioning model, except for the last row that summarizes the results for the learning model for arbitrary $k$ and $\ell$.
	}
	\label{tab:overview}
	\vspace{-7mm}
\end{table*}

\medskip
\noindent
\textbf{Algorithmic techniques.} The variant for general $k$ is still unresolved, hence it is vital to summarize algorithmic techniques used in this paper.
A straightforward analysis of the algorithm from Section~\ref{sec:k3} results in the bound $O(\ell^2)$, and the analysis must be improved in two places. One must bound the cost of each of $O(\ell)$ reconfigurations per~phase by a~constant, and show that the optimum offline algorithm must pay a~significant cost for inter-cluster requests.
We bound the cost of the latter by estimating the capabilities of the optimum offline algorithm to \emph{prepare} for an incoming sequence of requests.
Furthermore, in the analysis of the algorithm from Section~\ref{sec:k2}, we propose a~novel charging scheme for edges that share a~vertex.
We also emphasize the implications of the tight result for the learning model.
If one hopes to improve the lower bound for the general model, the ground sets must be dynamically added and \emph{removed}: if ground sets are only added, the instance is solvable by the algorithm from Section~\ref{sec:ppl} within the cost $O(k\cdot \ell)$.

\section{The Learning Model} %$\Omega(k\cdot \ell)$ for Competitive Ratio of Any Deterministic Algorithm}

In this section, we consider the learning variant of dynamic balanced graph partitioning problem.
For this setting, we show a~surprisingly high lower bound of $\Omega(k \cdot \ell)$ for $k\geq 3$.
The lower bound holds also in the general partitioning model (studied in Section~\ref{sec:part}).
At the end of this section, we discuss an asymptotically optimal upper bound for the learning variant.

\subsection{Lower Bound}

\label{sec:lowerbound}

We provide a~lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic online algorithm for the learning problem.
Later, we elaborate on how to efficiently transform it to a~lower bound for the general partitioning problem.
The lower bound requires $k\geq 3$.
In contrast, for $k=2$ the learning problem is trivial: immediate collocation of communicating pairs is $1$-competitive.
In contrast,
the general partitioning problem for $k=2$ is non-trivial (see Section~\ref{sec:k2}).
%, a~lower bound of $3$ is known~\cite{repartition-disc}, and we provide a~$6$-competitive algorithm, see Section~\ref{sec:k2}.)

Throughout this paper, we often refer to groups of communicating nodes.
We use this concept slightly differently in the lower bound than the upper bounds.
In our algorithms, we group nodes with a~communication history into \emph{components}.
In this section,
we group nodes that may ever communicate, into \emph{ground sets}.

Given a perfect partition,
every subset of nodes that belong to the same cluster in this partition
is a~\emph{ground set}. 
Any competitive algorithm under the learning model
 maintains a~perfect partition of ground sets into clusters.
 On each inter-cluster request, a ground set is revealed.
 An algorithm recovers the (hidden) perfect partition gradually over inter-cluster requests,
  by merging pairs of ground sets involved in these requests.

The adversary constructs ground sets depending on the choices of a deterministic online algorithm.
Once we construct a ground set, it lasts until the end of the input sequence.
We say that a ground set is a~\emph{singleton} if it contains exactly one node,
which is an \emph{isolated} node.

We start by constructing a~ground set of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$.
%, and we refer to it as the \emph{pivot} node.
We issue requests between this node and some node that was initially collocated with it.
By repeating such requests, almost every node is once collocated with the first ground set.
In comparison, we show that there exists an optimal offline algorithm \OPT that performs only two node exchanges ("swaps").
%Figure~\ref{fig:lb} illustrates the constructs used in the lower bound.

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for the learning model of Dynamic Balanced Graph Partitioning is at least
	 $(k-2)(\ell-1)/2 - 2$ for any $k\geq 3$ and $\ell \geq 2$.
\end{theorem}

\begin{proof}
	Fix any online algorithm \ALG.
	For a ground set $C$ of nodes that are initially collocated in one cluster,
	let $I(C)$ denote the cluster.
	We refer to $I(C)$ as the cluster of \emph{origin},
	when $C$ is clear from the context.
%	If $C$ contains nodes that originate from different clusters
%	then $I(C)$ is undefined.
%	Throughout the construction, we maintain ground sets for each node.
	Initially, all nodes are isolated, i.e., each node is in a singleton ground set.
	First, we choose a cluster arbitrarily and
	 create a~ground set $B$ of $k-1$ nodes in this cluster.
	%We never dismiss any constructed ground set.
	Each cluster hosts exactly $k$ nodes, and in any feasible partition,
	 a~single isolated node must be collocated with $B$.
	At any time,
	we refer to the isolated node currently collocated with $B$ as the \emph{pivot} node.
	Let $x_0$ denote the first pivot node.

	Then, we join the pivot node to a~larger ground set to force its eviction.
	Precisely,
	we create a~ground set $\{x_0, y_0\}$, 	where
%	 $x_0$ is the current pivot node and
	 $y_0$ is an arbitrary isolated node.
	Since \ALG does not have $\{x_0,y_0\}$ collocated, 
	the adversary issues an external request to this pair so that \ALG collocates it.
	\ALG cannot collocate $\{x_0, y_0\}$ with $B$ (as $B$'s size is $k-1$), hence it collocates them in a~different cluster.
	In order to preserve a~feasible partition of nodes after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another isolated node that becomes the new pivot.
%	denoted by $x_1$.

	We proceed in similar steps by joining the current pivot node to a ground set of the same origin residing in a different cluster.
	Consider the step $i$,
	when the isolated node $x_i$ is collocated with $B$.
	 We issue a request between $x_i$ and some node in $C_i$,
	 where $C_i$ is the largest ground set
	 s.t.~$I(C_i) = I(x_i), C_i \neq \{x_0,y_0\}$.
	 Then \ALG must collocate the new ground set
	  $\{x_i\} \cup C_i $ in one cluster.
%	 only if  a feasible partition exists
%	 that collocates all nodes in $\{x_i\} \cup C_i $.
	Any feasible partition replaces $x_i$ with some isolated node $x_{i+1}$,
	as the new ground set $\{x_i\} \cup C_i $  may not be ever split.
%	Otherwise,
%	no such feasible partition exists and
%	we terminate the process.
	We terminate the process once the number of remaining isolated nodes 
	is less than $\ell+3$.
	At each step $i$,
	the number of isolated nodes decreases either by one
	 or by two if $C_i$ is a singleton.
	Therefore, once the process terminates,
	in any case at least $\ell+1$ isolated nodes are left.

	  Next, we argue that a feasible partition exists when the process terminates.
	 This implies that a feasible partition exists after any earlier step as well.
%	 Assume for contradiction that the partition is not feasible.	 
	 Since there are at least $\ell+1$ isolated nodes left,
	 there must be two isolated nodes $x^*$ and $y^*$,
	  with the same cluster of origin,
	 i.e., $I(\{x^*\}) = I(\{y^*\})$.
	 Consider the partition $P^*$
	  obtained from the initial partition after swapping $x_0$ and $y_0$
	 with $x^*$ and $y^*$ (respectively).
	 In this partition,
	 the ground set $\{x_0,y_0\}$ is collocated in the cluster $I(\{x^*,y^*\})$.
	 Note that after the first request $\{x_0,y_0\}$,
	   we issue requests only between nodes that have the same cluster of origin and
	   all these nodes are collocated in $P^*$.
	 Therefore all ground sets constructed so far are collocated in  $P^*$,
	 and it is a feasible partition.
%	 Hence, there are at most $\ell$ isolated nodes after the termination.

%	The number of isolated nodes decrease after each step.
%	Consider the last step, $j$, when  $\ell+1$ isolated nodes exist before  $x_{j}$ joins to another ground set.
%	That is, there are less than $\ell+1$ isolated nodes after the step $j$.
%	Then, at the end of step $j-1$,
%	there exist two isolated nodes $x^*$ and $y^*$
%	with the same cluster of origin, i.e., $I(\{x^*\})=I(\{y^*\})$.
%
%   Next,
%    we bound costs incurred by \OPT and \ALG over the  sequence of requests
%    $\sigma := \langle \{x_0,y_0\}, \dots, \{x_{j-1},y_{j-1}\}  \rangle$.

	Consider  nodes $x^*$ and $y^*$ and the partition $P^*$ obtained previously.
	\OPT moves to $P^*$ by performing only two node swaps.
	Precisely, \OPT collocates $\set{x_0,y_0}$ 
	by swapping them with $x^*$ and $y^*$.
	No ground set is split in $P^*$ and
	 \OPT pays only for the two swaps.

	\ALG performs at least one swap at each step $i$,
	and some ground set grows.
%	In other words,
%	some ground set grows by one node at each of these steps.
	Consider any ground set $C^* \neq B$ after the termination.
	This ground set has grown exactly $|C^*|-1$ times until the termination. 
	Let $\mathcal{S}$ be the set of all ground sets after the process terminates.
	Thus, $\mathcal{S}$ includes ground sets $B$,
	$\{x_0,y_0\}$,
	and (up to) $\ell+2$ singleton ground sets.
	Among the remaining ground sets in $\mathcal{S}$,
	no two ground sets have the same origin.
	Otherwise,
   the smaller ground set is either a singleton,
   which contradicts the bound $\ell+2$ on the number of singletons,
	or we have joined nodes to it at some step,
	contradicting our choice of the largest~$C_i$ at step $i$.
	Hence,  there are at most $\ell-1$ such ground sets,
	one per possible cluster of origin,
	 excluding the cluster containing $B$. 
	Therefore,  $|\mathcal{S}| \leq 1 + 1 + (\ell+2)  + (\ell-1) = 2\ell+3$.
	Note that among all non-singleton ground sets in $\mathcal{S}$,
	 only $B$ does not grow during the process.
	 Thus,
    the total number of times that 
	a ground set in $\mathcal{S}$  has grown is
	$	\sum_{C^* \in \mathcal{S} } (|C^*|-1) - (k-1) $
	\begin{align*}
	=\sum_{C^* \in \mathcal{S} } |C^*| - 
		\sum_{C^* \in \mathcal{S} } 1 - (k-1) 
	\geq  k  \ell - (2\ell+3) - (k-1) =  (k-2)(\ell-1) - 4,
	\end{align*}
	which bounds the number of swaps performed by \ALG.
	The competitive ratio is then
	$\ALG / \OPT \geq ((k-2)(\ell-1) - 4) / 2$.
	%%
	\begin{comment}
	\ALG performs at least one swap  every time   a~ground set is constructed.
	Each expansion reduces the number of isolated nodes by $1$ or $2$.
	The decrease of two isolated nodes occurs when the largest component on  $I(x_i)$ is still singleton; this occurs at most $\ell$ times.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	Hence, the total number of swaps of \ALG is at least $k\cdot \ell - k - 2 \ell$.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - 2\ell) / 2$.
	\end{comment}
	%%
\end{proof}

\noindent
\textbf{Lower bound for the general problem.}
For a~lower bound $\Omega(k\cdot \ell)$ for the general partitioning problem, the adversary continues to issue requests to split nodes of a~ground set until the algorithm collocates them.
Note that the ground sets constructed in our lower bound can be perfectly partitioned into the clusters.
Hence, the optimal algorithm moves to a~perfect partition at the beginning (where requests incur the cost~$0$), and its cost is bounded.
This means that the algorithm must eventually collocate all nodes of a~ground set to be competitive.
We reveal the next ground set only after the collocation, hence we can repeat the analysis of the algorithm for the learning problem.
Finally, we note that the construction is oblivious to the choice of the reconfiguration cost $\alpha$: we compare the 	number of node exchanges of \ALG and \OPT.

\medskip

\noindent
\textbf{Resource augmentation.}
The majority of work on the online balanced partitioning problem so far~\cite{repartition-disc,sigmetrics19_partitioning} focuses on the scenario with resource augmentation, where the clusters of an online algorithm are larger than the clusters of the offline optimal algorithm that we compare the performance to.
We can adjust our construction to show a~lower bound of $\Omega(\ell)$ for a~setting with resource augmentation.

Consider a~partitioning problem with resource augmentation $1+1/3-\epsilon$.
Fix $k$ divisible by $3$, and construct $3$ ground sets of size $k/3$ in each cluster.
Note that no more than $3$ such ground sets fit in one cluster.
Then, apply the construction from the lower bound for $k=3$, using these ground sets in the way we used individual nodes.
The cost of any algorithm (including \OPT) scales up by $k/3$, and the lower bound $\Omega(\ell)$ holds.

Finally, we note the possibility of improvement. The algorithm CREP~\cite{repartition-disc} requires $(2+\epsilon)$-augmentation to guarantee the competitive ratio independent of $\ell$.
In contrast, our construction shows that the linear term $\ell$ is inevitable if the augmentation is smaller than~$1+1/3$.

\subsection{Upper Bound}
\label{sec:ppl}

We present an asymptotically optimal algorithm for the learning problem.
The algorithm  collocates  a~pair as soon as they communicate and it never separates them.
In order to preserve collocated pairs,
we employ the concept of components,
introduced by Avin~et~al.~\cite{repartition-disc}.

We maintain subsets of frequently communicating nodes as \emph{components}.
Initially,
each node constitutes a single-node component which we refer to as a \emph{singleton} component, 
and the node in such component is an \emph{isolated} node.
We define larger components in terms of smaller components.
Concretely,
given a (sub)sequence of requests,
two components $C_1$ and $C_2$
 \emph{merge} into one component as soon as
 for some pair of nodes $v_1 \in C_1$ and $v_2 \in C_2$,
the frequency of requests $\{v_1,v_2\}$ reaches a certain threshold through the sequence.
We keep all nodes of a~component always collocated in the same cluster,
i.e., when we  move a~node,
we move the whole component that contains it.
A~partition that has every component collocated 
is a~\emph{component respecting} partition.
%Intuitively, a component preserves the history of requests between its nodes.

In addition,
we maintain a~balanced partition of our components as long as such partition exists,
a reminiscent of partitioning given integers into sets of equal sum~\cite{integer-partitions-book}.
In contrast, our partition is time-varying:  two components are merged into one component once they communicate, and we adjust the partition accordingly.

The algorithm in this section and the algorithm for $k=3$ (cf. Section~\ref{sec:k3}) are modified versions of the algorithm DET from~\cite{repartition-disc}.
The difference is in the choice of partition after a~component merge.
In DET, the partition was arbitrary.
In the algorithm for the learning model,
we choose a~component respecting partition closest to the initial partition.
In the algorithm from Section~\ref{sec:k3}, we choose the partition closest to the current partition (a~repartition of minimum cost).
%A generic sketch of these two algorithms can be found in the appendix (Algorithm \ref{alg:ppl}).

\noindent
\textbf{Perfect Partition Learner algorithm.}
Now we describe the algorithm \PPL.
On each inter-cluster request $\{u,v\}$, 
\PPL creates new components by merging
 the two components that contain nodes $u$ and $v$.
In order to collocate nodes of the new component,
\PPL moves to a~component respecting partition that minimizes the distance to the initial partition $P_I$.
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

Fix the initial partition
$P_I :=\{ I_1, \dots, I_{\ell}\}$ and \OPT's final partition
$P_F := \{F_1, \dots, F_{\ell}\}$.
The \emph{distance} of a~partition
 $P = \{C_1, \dots, C_{\ell}\}$ from the initial partition,
defined as 
$\Delta(P) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$,
 is the number of nodes in $P$ that do not reside in their initial cluster.
In other words,
at least $\Delta(P)$ node migrations are required in order to reach the partition $P$ from $P_I$, and thus
$\OPT \geq \Delta(P_F) $.

\PPL never moves to a~partition that is more than $ \Delta^*:= \Delta(P_F) $   migrations away from~$P_I$.
This invariant latter ensures us that \PPL does not pay too much while recovering $P_F$.
We emphasize that a~\emph{repartitioning} by \PPL replaces the current partition $P$ with a~perfect partition closest to $P_I$.
This way \PPL never moves to a~partition beyond the distance $\Delta^*$.

\begin{property} \label{prop:dist<OPT}
	Let $P$ be any partition chosen by \PPL at any time.
	Then, $\Delta(P) \leq \Delta^*$.	
\end{property}

\begin{lemma}	\label{lemma:rebalancecost}
	The cost of each repartitioning by \PPL is $2\cdot\OPT$.
\end{lemma}
\begin{proof}
	Let $P_i$ denote the partition of \PPL immediately after serving $\sigma_{i}$.
	Consider the repartitioning that transforms 
	$P_{t-1}$ to $P_t$ upon the request $\sigma_t$.
	Let $M \subset V$ denote the set of nodes that migrate during this process.
	Let $M^-$ and $M^+$ denote the subset of nodes that, respectively,
	enter or leave their initial cluster during the repartitioning.    
	Then,
	$M = M^+ \cup M^-$.
	Since at least $|M^-|$ nodes are not in their initial cluster before the repartitioning (i.e., in $P_{t-1}$),
	the distance before the repartitioning is $\Delta(P_{t-1}) \geq | M^-|$.
	Analogously,
	the distance afterward is $\Delta(P_{t}) \geq | M^+|$.
	Thus,
	$|M| \leq \Delta(P_{t-1}) + \Delta(P_{t})$.
	By Property \ref{prop:dist<OPT},
	$\Delta(P_{t-1})  \leq \Delta^*$ and
	$\Delta(P_{t}) \leq \Delta^*$.
	Since $\Delta^* \leq \OPT$,
	we obtain	 $|M| \leq 2\cdot\OPT$.
\end{proof}

>>>>

\documentclass{standalone}
\usepackage{tikz}
 \usetikzlibrary{decorations.pathreplacing} %

\begin{document}
		\begin{tikzpicture}
	\draw [->] (-10mm,0) -- (5,0);
	\draw [dotted] (-13mm,0) -- (-10mm,0);
